---
title: "Garrett's Final"
author: "Garrett Weems"
date: "2024-04-23"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Mediterranean Diet 
Garrett Weems
2024

## Introduction
A trip to the grocery store may be a worthwhile experience; you are preparing to do something that will save you money and provide nutrition for you and your loved ones. You consider healthy eating but are confused by the absurd amount of recommendations you have heard through the media. "The basic principles of good diets are so simple that I can summarize them in just ten words: eat less, move more, eat lots of fruits and vegetables." It seems that the Mediterranean diet fits the bill. 

### My Approach
Is there any correlation between certain foods eaten and heart health? First, determine the eating patterns of people who follow the Mediterranean diet, especially looking at foods of high interest when regarding health such as flavanoid rich foods and vegetables.   

## Data

In 2020, researchers Amato et. al investigated the eating patterns of people who had heart problems and compared them with those without. The data includes how often these participants (mostly men) ate foods such as canned fish, specific fruits and vegetables, simple pastas, etc. 

```{r, echo = FALSE, warning= FALSE, message = FALSE, results= 'hide'}
library("dplyr")#needed so I can mutate with foods consumed with a total fruits and vegetables amount. 
library("tidyr")#since each observation is a row
library("ggplot2")#useful for my plotting of the data 
library("readxl")#needed to read the excel files 
library("vctrs")
suppressWarnings({ 
}) 
```

## What is the BMI for the Participants? 

```{r, echo = FALSE, warning= FALSE, results='hide', include=FALSE}
#create my own copies of the dataset
clinical_data <- read_xlsx("clinical_data.xlsx")
ffq_data <- read_xls("FFQ_food_frequencies.xls")
grams_data <- read_xls("food_grams.xls")

#look at the clinical data to determine health of the participants
bmi_col <- clinical_data[,"BMI" ]
summary(bmi_col)
mean_bmi <- sapply(bmi_col, mean, na.rm = TRUE)
rounded_mean_bmi <- signif(mean_bmi, 3)


```

The mean BMI for these older men from Italy is `r rounded_mean_bmi`, which is, if anything, slightly overweright yet normal according to standards according to statistics published on statista [here](https://www.statista.com/statistics/727866/distribution-of-body-mass-index-by-age-italy/) but be quick since you are here for MY data and graphs! 




## The Plot Thickens 
```{r,include=FALSE}
#Do a graph of the BMI by age group 
#Create age groups and see if that is similar to the statista data - does BMI steaily increase with age? 
#create age groups by grabbing the rows of data that have people aged x-y, y-z, z-a, etc.
age_col <- clinical_data[, "Age"]
min_age <- sapply(age_col, min, na.rm = TRUE) %>%  signif(2)
max_age <- sapply(age_col, max, na.rm = TRUE) %>%  signif(2)

med_age <- sapply(age_col, median, na.rm = TRUE)  %>%  signif(2)
colnames(clinical_data)

#simply do by what deca(de) of life they are in
thirty_bmi <- unlist(clinical_data %>%  filter(Age %>%  between(30,39)) %>% select(BMI, Age))
forty_bmi <- unlist(clinical_data %>%  filter(Age %>%  between(40,49)) %>% select(BMI, Age))
fifty_bmi <- unlist(clinical_data %>% filter(Age %>%  between(50,59)) %>% select(BMI, Age))
sixty_bmi <- unlist(clinical_data %>%  filter(Age %>%  between(60,69)) %>% select(BMI, Age))
seventy_bmi <- unlist(clinical_data %>% filter(Age %>%  between(70,79)) %>% select(BMI, Age))
class(thirty_bmi)
#now data is in multiple tibbles. Next I need to get them all in one tall dataframe in order to plot them groupby
combined_data <- bind_rows(thirty_bmi, forty_bmi, fifty_bmi, sixty_bmi, seventy_bmi)

```
That is interesting because we are told that the Mediterranean diet leads to great health. Is BMI not that much of a factor in how long you live then? Men being tested are anywhere from `r min_age` to `r max_age` years old. Let's continue to get a graph of how different age groups' BMI varies. 


